---
title: "Homework 04: Tidy data and joins"
author: "Stephen Chignell"
date: "October 9, 2018"
output:
  github_document:
    toc: yes
---

## Overview

**Goal:** Solidify data wrangling skills by working on some realistic problems in the grey area between data aggregation and data reshaping.

**Assignment:** This is a “choose your own adventure”-style assignment, where you are expected to do the following two things:

- Pick one of the data reshaping prompts and do it.
- Pick one of the join prompts and do it.

It is fine to work with a new dataset and/or create variations on these problem themes.

## Reshaping Activity #1: Cheat Sheet

I have developed a separate reshaping [Cheat Sheet](https://github.com/STAT545-UBC-students/hw04-schignel/blob/master/Reshaping_in_tidyR_-_cheat_sheet.md), which is stored in my Homework 4 repository.


## Joining Activity #1: Supplementary data

**Instructions:** Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result.

####Load libraries
```{r gapminder}
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(gapminder))
```

####Create new table
Next, we will create a new data frame with supplementary information to join to`gapminder`.

```{r supplementary}
# Create a list of country names
country <- c("Argentina", "Ethiopia", "Senegal", "Indonesia", "China", "Canada", "Germany", "Australia", "South Sudan") 

# List area (km2) for each country
area_km2 <- c(2780400, 1127127, 196190, 1919440, 9706961, 9984670, 357021, 7692024, 619745)

# List capital city for each country
capital <- c("Buenos Aires", "Addis Ababa", "Dakar", "Jakarta", "Beijing", "Ottowa", "Berlin", "Canberra", "Juba")

#Save vectors to new data frame
supp <- data.frame(country, area_km2, capital)

#View supplementary data frame as a nice table
knitr::kable(supp, caption = "Supplementary country data")
```

**Supplementary data sources**

+ [Country areas](https://simple.wikipedia.org/wiki/List_of_countries_by_area?oldformat=true)
+ [Country capitals](https://www.countries-ofthe-world.com/capitals-of-the-world.html)


We are now ready to practice different types of joins between the supplementatry data frame and the gapminder dataset. 

####Left Join
`left_join` is the most common way to merge to data sets:
```{r left join, warning=FALSE}
#Left join and pipe to str() to inspect without generating massive table
left_join(gapminder, supp, by = "country") %>% 
  str() 
```

Because we specified `gapminder` as the lefthand column, `left_join` kept all of the `gapminder` records while successfully joining the `area_km2` and `capital` variables to it. Notice that the funtion has inserted "NA" values for countries not listed in the supplementary table. 

####Right Join
`right_join` is the same as `left_join`, except that it returns all records from the righthand table (while retaining all columns from both tables).
```{r right join, warning=FALSE}
#Right join and pipe to str()
right_join(gapminder, supp, by = "country") %>% 
  str()
```
Here we see that the `gapminder` dataset has been reduced from 1704 to 97 observations, and that the two additional variables from the supplementary table have been joined.

####Inner Join
What if we wanted an output containing only the countries listed in both datasets?

For this we use the `inner_join` function:
```{r inner join, warning=FALSE}
#Inner join and nice table to preview
knitr::kable(head(inner <- inner_join(gapminder, supp, by = "country")))
```

We can see that the variables `area_km2` and `capital` are joined, but we can't see from this table whether or not it removed the records that didn't match. Let's check with the `str()` function.  
```{r inner str}
#Look at the structure
str(inner)
```

Just as we would expect `gapminder` has been reduced from 1704 to 96 observations, and the two additional variables from the supplementary table have been joined. 

**Question** Why does the `inner_join` output have 96 observations while the `right_join` output had 97? 

####Anti Join
We can find the source of this discrepancy using an `anti_join`, which returns all rows from the first table where there are not matching values in the second, filtering the output to include just columns from the first table.
```{r anti join, warning=FALSE}
# Supplementary table is first listed to identify discrepancy
knitr::kable(anti_join(supp, gapminder, by = "country"))
```

Interpretating this result, we can recall that South Sudan became a country in 2011, and is therefore not included in the `gapminder` data set, which includes data up to 2007.

####Full Join
To keep all data in both tables, we can use the `full_join` function
```{r full join, warning=FALSE}
str((full_join(gapminder, supp, by = "country")))
```
Here we see that the output has 1705 observations of 8 variables, indicating that operation has kept all data from both tables, and replaced missing data cells with "NA".



