---
title: "Reshaping in tidyr (cheat sheet)"
author: "Stephen Chignell"
date: "October 9, 2018"
output:
  github_document:
    toc: yes
---

##Reshaping data

One of the strengths of `dplyr` and `tidyr` is their ability to quickly change the layout of "messy" data into data structures that are conducive for exploratory analysis and graphing. 

###Functions
The primary functions are below. 

| Function       | What it does                     | Another way to think about it   |
|----------------|----------------------------------|---------------------------------|
|  `gather()`    | transform data from wide to long | Gather columns into rows        |
|  `spread()`    | transform data from long to wide | Spread rows into columns        |
|  `separate()`  | split 1 variable into multiple   | Separate 1 column into several  |
|  `unite()`     | merge multiple varialbes into 1  | Unite several columns into 1    |


- "Wide" = More **Horizontal**
- "Long" = More **Vertical**


**Arguments**

`gather(data, key, value, na.rm = FALSE)`

+ `data` = the data frame (can omit when piping)
+ `key` = what you want to call the new column name from headers
+ `value` = what you want to call the new column name from stacked columns 
+ `na.rm` = items to include/exclude
    + Use a "-" to speficify an exclusion
    + You can specify single or multiple items
    + If left blank, it will try to gather **all** columns
    


###Demonstration
We will demo the `spread` and `gather`, since they are the most commonly used. First, load the required libraries:

```{r libraries}
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(knitr))

```

####Create some sample data
Suppose you have some data on baseball teams and their runs in a number of games:

```{r baseball}
baseball <- data.frame(
   ID = LETTERS[1:10],
   Blue_Jays = sample(0:1, 5, TRUE),
   Cardinals = sample(0:1, 5, TRUE),
   Cubs = sample(0:1, 5, TRUE),
   Mariners = sample(0:1, 5, TRUE)
   )
knitr::kable(baseball)
```

####Gather the data
```{r gather}
bb.long <- gather(baseball,Teams, Runs, -c(ID)) #exclude the "ID" field
knitr::kable(bb.long)
```


Notice it is now in **long** format, and "ID" has been excluded from the `gather` process.


####Spread the data back out
```{r spread}
bb.long %>% 
  spread(Teams, Runs) %>% #no need to specify dataset when piping
  knitr::kable()
```

###Additional Resources
In order of increasing detail:

+ [tidyr in a nutshell](https://github.com/trinker/tidyr_in_a_nutshell/blob/master/README.md)
+ [RStudio Data Wrangling (Ctrl+F for "gather")](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
+ [Guru99 dplyr](https://www.guru99.com/r-dplyr-tutorial.html)
+ [tidyverse documentation](https://tidyr.tidyverse.org/index.html)
